---
title: "mimic_heart"
author: "Sharon Lee"
date: "5/13/2020"
output: html_document
---
Exploration and Modeling of a MIMIC III Cohort:
Looking at premature heart attack (65 years and younger) and factors in order to predict outcome (mortality).

```{r setup, include=FALSE}
library(ggplot2)
library(caret)
library(rpart.plot)
library(pROC)
library(randomForest)
library("pwr")
library(corrplot)
library("rgl")
library("pscl")
library("car")
options(warn=-1)
```

#read in our dataset
```{r}
setwd('C:/USFMSHI/HS 614/MIMIC')
heart <- read.csv("data_heartattack_final1.csv", header = T)
```

################################################### EDA ##########################################################
```{r}
summary(heart)
```

#in total, we have 1835 patients in this dataset
#drop X
#icd9_list
#drop height (1456 NAs) 
#drop weight (1293 NAs)
#drop total_cholesterol  (961 NAs)
#drop hdl_cholesterol (977 NAs)
#drop ldl_cholesterol  (1018 NAs)
#drop triglycerides  (935 NAs)
#drop crp (1757 NAs)   
#drop hba1c  (1170 NAs) 
#drop mean_bp (1546 NAs) 
#drop heart_rate (1293 NAs)    
#drop max_resp_rate  (1293 NAs)         
#drop procedure_codes 
```{r}
heart <- heart[ , -c(1, 10, 11, 12, 14, 15, 16, 17, 20, 23, 25, 27, 28, 29)]

#we don't need hadm_id, subject_id, diagnosis, discharge location as well
heart <- heart[ , -c(1, 2, 5, 6)]
summary(heart)
```

#remove rows with more than 3NAs
```{r}
heart <- heart[-which(rowSums(is.na(heart)) >= 3),]
```

#ethnicity
```{r}
summary(heart$ethnicity)  #there are 24 ethnicity types in this dataset 

#regroup ethnicity types: asian, black, hispanic, white, others 
lev_e <- c("ASIAN", "BLACK", "HISPANIC", "WHITE", "OTHERS")
heart$ethnicities <- factor(rep(NA, nrow(heart)), order = F, levels = lev_e)
heart$ethnicities[heart$ethnicity == "PATIENT DECLINED TO ANSWER" | 
                  heart$ethnicity == "UNABLE TO OBTAIN"|
                  heart$ethnicity == "UNKNOWN/NOT SPECIFIED"] <- NA
heart$ethnicities[heart$ethnicity == "ASIAN"| 
                  heart$ethnicity == "ASIAN - ASIAN INDIAN" |
                  heart$ethnicity == "ASIAN - CAMBODIAN" |
                  heart$ethnicity == "ASIAN - CHINESE" |
                  heart$ethnicity == "ASIAN - VIETNAMESE"] <- "ASIAN"

heart$ethnicities[heart$ethnicity == "BLACK/AFRICAN AMERICAN"| 
                    heart$ethnicity == "BLACK/CAPE VERDEAN" |
                    heart$ethnicity == "BLACK/HAITIAN"] <- "BLACK"

heart$ethnicities[heart$ethnicity == "HISPANIC/LATINO - CUBAN" |
                    heart$ethnicity == "HISPANIC OR LATINO" |
                    heart$ethnicity == "HISPANIC/LATINO - CUBAN"] <- "HISPANIC"

heart$ethnicities[heart$ethnicity == "PORTUGUESE" |
                    heart$ethnicity == "WHITE" |
                    heart$ethnicity == "WHITE - BRAZILIAN" |
                    heart$ethnicity == "WHITE - OTHER EUROPEAN" |
                    heart$ethnicity == "WHITE - RUSSIAN"] <- "WHITE"

heart$ethnicities[heart$ethnicity == "MIDDLE EASTERN" |
                    heart$ethnicity == "MULTI RACE ETHNICITY" |
                    heart$ethnicity == "NATIVE HAWAIIAN OR OTHER PACIFIC ISLANDER" |
                    heart$ethnicity == "OTHER"] <- "OTHERS"
```

```{r}
#make sure new factor has the correct count for each level
summary(heart$ethnicities) #345NAs
#remove ethniity column
heart <- heart[ ,-1]
#remove rows with NAs
heart <- heart[-which(is.na(heart$ethnicities)),]
```

#expire_flag
```{r}
#change to factor 
class(heart$expire_flag) #right now is int
heart$expire_flag <- as.factor(heart$expire_flag)
levels(heart$expire_flag) <- c("no", "yes")
summary(heart$expire_flag) #make sure there are no NAs
```

#age 
```{r}
summary(heart$age) #we want premature heart attack, making sure maximum age is 65
```

#iculos (in fractional days)
```{r}
summary(heart$iculos)
#2NA's
heart[which(is.na(heart$iculos)), ]

#impute with mean 
icu_mean <- mean(heart$iculos, na.rm = TRUE)
heart$iculos[is.na(heart$iculos)] <- icu_mean
```

#red_blood_cells
```{r}
class(heart$red_blood_cells) #factor right now, change to numeric

heart$red_blood_cells[heart$red_blood_cells == 'UNABLE TO REPORT'] <- NA
heart$red_blood_cells[heart$red_blood_cells == 'UNABLE TO REPORT DUE MCHC >38.0'] <- NA
heart$red_blood_cells <- as.numeric(as.character(heart$red_blood_cells))
summary(heart$red_blood_cells) 

heart[which(is.na(heart$red_blood_cells)), ] #impute with mean

rbc_mean <- mean(heart$red_blood_cells, na.rm = TRUE)
heart$red_blood_cells[is.na(heart$red_blood_cells)] <- rbc_mean
summary(heart$red_blood_cells)
```

#hb
```{r}
summary(heart$hb)  #371 NAs
#drop hb column
heart <- heart[,-6]
```

#creatinine kinase need to be changed to numeric
```{r}
heart$creatinine_kinase[heart$creatinine_kinase == "ERROR"] <- NA
heart$creatinine_kinase <- as.numeric(as.character(heart$creatinine_kinase))
summary(heart$creatinine_kinase)  #66 NAs, min at 8 and max at 96400

heart[which(is.na(heart$creatinine_kinase)),]  
#impute with median
kinase_median <- median(heart$creatinine_kinase, na.rm = TRUE)
heart$creatinine_kinase[is.na(heart$creatinine_kinase)] <- kinase_median
```

#ck_isoenzyme
```{r}
heart$ck_isoenzyme  #NA, NotDone,Greater than 500  
heart$ck_isoenzyme[heart$ck_isoenzyme == 'NotDone'] <- NA
#substitue greater than 500 to NA
heart$ck_isoenzyme[heart$ck_isoenzyme == "GREATER THAN 500"] <- NA
#change to numeric
heart$ck_isoenzyme <- as.numeric(as.character(heart$ck_isoenzyme))
summary(heart$ck_isoenzyme)  #411 NAs
heart <- heart[,-7]
```

#glucose
```{r}
summary(heart$glucose)  #more than 445 NAs
#drop glucose
heart <- heart[,-7]
```

#heart rhythm
```{r}
summary(heart$heart_rhythm)  #236 NAs 
heart$heart_rhythm <- droplevels(heart$heart_rhythm)

#remove rows with NA
heart <- heart[-which(is.na(heart$heart_rhythm)), ]
```

#data cleaning complete
```{r}
summary(heart)
```

correlations 
#create new dataframe that holds only numeric variables 
```{r}
heart_num <- heart[, -c(1, 2, 7, 8)]

#change all factor variables to numeric 
heart_num$gender <- as.numeric(heart$gender)
heart_num$expire_flag <- as.numeric(heart$expire_flag)
heart_num$heart_rhythm <- as.numeric(heart$heart_rhythm)
heart_num$ethnicities <- as.numeric(heart$ethnicities)

#use spearman method when calculate correlation since our dataframe now has ranks 
heart_cor <- cor(heart_num, use = "pairwise.complete.obs", method = "spearman")
heart_cor
corrplot(heart_cor)
```

############################################## Data Visualizations ###############################################
#ethnicities
```{r}
g1 <- ggplot(heart, aes(ethnicities))
g1 + geom_bar(aes(fill = ethnicities), width = 0.75) +
  theme(axis.text.x = element_text(angle = 75, vjust=0.6)) +
  labs(title = "Ethnicity Distribution in Heart Dataset")
```

#gender
```{r}
g2 <- ggplot(heart, aes(gender))
g2 + geom_bar(aes(fill = gender), width = 0.75)  + 
  labs(title="Gender Distribution in Heart Dataset") 
```

#expire flag
```{r}
g3 <- ggplot(heart, aes(expire_flag))
g3 + geom_bar(aes(fill = gender), width = 0.75)  + 
  labs(title="Expire_flag Distribution in Heart Dataset colored by Gender") 
```

#age
```{r}
g4 <- ggplot(heart, aes(age))
g4 + geom_histogram(aes(col = gender), bins = 30) + 
  labs(title="Age Distribution in Heart Dataset colored by Gender") 
```

#iculos
```{r}
g5 <- ggplot(heart, aes(iculos))
g5 + geom_histogram(aes(fill = expire_flag), bins = 30) + 
  labs(title="ICU Length of Stay Distribution in Heart Dataset colored by Expire_flag") 
```

#red blood cells
```{r}
g6 <- ggplot(heart, aes(red_blood_cells))
g6 + geom_histogram(aes(fill = gender), bins = 30) + 
  labs(title="ICU Length of Stay Distribution in Heart Dataset colored by Gender") 
```

#creatine kinase
```{r}
g7 <- ggplot(heart, aes(creatinine_kinase))
g7 + geom_histogram(aes(fill = gender)) + 
  labs(title="Creatine kinase Distribution in Heart Dataset colored by Gender") 
```

#heart_rhythm 
```{r}
g8 <- ggplot(heart, aes(heart_rhythm))
g8 + geom_bar(aes(fill = heart_rhythm)) +
  labs(title = "Heart rhythm distribution types")
```

#heart rhythm and expire flag
```{r}
g8 + geom_bar(aes(fill = expire_flag))
```

#iculos and heart heart rhythm and gender 
```{r}
g9 <- ggplot(heart, aes(heart_rhythm, iculos))
g9 + geom_boxplot(aes(fill = gender))
```

#iculos and age 
```{r}
g10 <- ggplot(heart, aes(age, creatinine_kinase))
g10 + geom_point(aes(col = gender)) + ylim(0, 10000)
```

#expire flag and ethnicity
```{r}
g3 + geom_bar(aes(fill = ethnicities), width = 0.75)  + 
  labs(title="Expire_flag Distribution in Heart Dataset colored by Ethnicities") 
```

################################################### Modeling ####################################################

#Split your data into train and test sets
```{r}
set.seed(3233)
intrain <- createDataPartition(y = heart$expire_flag, p= 0.7, list = FALSE)
training <- heart[intrain,]
testing <- heart[-intrain,]
dim(intrain); dim(training); dim(testing)
```

#check into preprocessing
```{r}
anyNA(heart)
summary(heart)
```

#logistic regression model
```{r}
glmfit_all <- glm(expire_flag ~ ., family = binomial(),data = training)
summary(glmfit_all)
pR2(glmfit_all)
```

#use step function to find the best logistic regression model
```{r}
fit_null <- glm(expire_flag ~ 1, data = training, family = binomial())
fit_step = step(fit_null, scope=list(lower=fit_null, upper=glmfit_all),direction="both")
summary(fit_step)
pR2(fit_step)
```

#predict with log model
```{r}
p <- predict(fit_step, testing, type = "response")
p
```

#confusion matrix
```{r}
expire.pred = rep("no", dim(testing)[1])
expire.pred[p > 0.5] = "yes"
confusionMatrix(factor(testing$expire_flag), factor(expire.pred), positive = "yes")  
```

#svm classifier
#linear
```{r}
trctrl <- trainControl(method = "repeatedcv", number = 10, repeats = 3)
grid <- expand.grid(C = seq(0, 2, length =20))
set.seed(3233)

#Training peformed using either formula interface or matrix
svm_Linear_Grid <- train(expire_flag ~., data = training, method = "svmLinear",
                         trControl=trctrl,
                         preProcess = c("center", "scale"),
                         tuneGrid = grid,
                         tuneLength = 10) 
svm_Linear_Grid 
plot(svm_Linear_Grid)
```

#using svm_linear_Grid to predict on testing set 
```{r}
test_pred <- predict(svm_Linear_Grid, newdata = testing)
test_pred
```

#confusion matrix
```{r}
confusionMatrix(factor(testing$expire_flag), test_pred, positive="yes")
```

#poly
```{r}
svm_poly_Grid <- train(expire_flag ~., data = training, method = "svmPoly",
                         trControl=trctrl,                 
                         preProcess = c("scale","center"),
                         tuneLength = 4)
svm_poly_Grid
plot(svm_poly_Grid)
```

#using svm_poly_grid to predict on testing set 
```{r}
test_pred <- predict(svm_poly_Grid, newdata = testing)
test_pred
```

#confusion matrix
```{r}
confusionMatrix(factor(testing$expire_flag), test_pred, positive="yes") #65%
```

#rbf
```{r}
trctrl_rbf <- trainControl(method = "repeatedcv", number = 10, repeats = 10)
svmGrid <- expand.grid(sigma = c(0,0.01, 0.02, 0.025, 0.03, 0.04,
                                 0.05, 0.06, 0.07,0.08, 0.09, 0.1, 0.25, 0.5, 0.75,0.9),
                       C = c(0,0.01, 0.05, 0.1, 0.25, 0.5, 0.75,
                             1, 1.5, 2,5))
set.seed(3233)
svm_Radial <- train(expire_flag ~., data = training, method = "svmRadial",
                    trControl=trctrl_rbf,
                    preProcess = c("center", "scale"),
                    tuneGrid = svmGrid,
                    tuneLength = 10) #param grid 10 values C selected by caret
svm_Radial
plot(svm_Radial)
```

#using svm_Radial to predict on testing test
```{r}
test_pred <- predict(svm_Radial, newdata = testing)
test_pred
```

#confusion matrix
```{r}
confusionMatrix(testing$expire_flag, test_pred, positive = "yes")   
```

#Decision tree
```{r}
trctrl <- trainControl(summaryFunction=twoClassSummary,classProbs = TRUE,# Use AUC to pick the best model
                       savePredictions = T, method = "cv", number = 5)
set.seed(3233)
dtree_fit <- train(expire_flag ~., data = training, method = "rpart",
                   # selects splits, default is gini (impurity)
                   parms = list(split = "information"), #information gain
                   trControl=trctrl,  # same as above
                   tuneLength = 10)
dtree_fit
```

#using dtree_fit to predict on testing set 
```{r}
test_pred <- predict(dtree_fit, newdata = testing)
test_pred


```

#confusion matrix
```{r}
confusionMatrix(factor(testing$expire_flag), test_pred, positive="yes")
```

#decision tree
```{r}
prp(dtree_fit$finalModel, box.palette = "Reds")
```

#Alt with default control
```{r}
set.seed(3233)

dtree_fit2 <- train(expire_flag ~., data = training, method = "rpart")
dtree_fit2
```

#using dtree_fit2 to predict on testing set 
```{r}
test_pred2 <- predict(dtree_fit2, newdata = testing)
test_pred2
```

#confusion matrix
```{r}
confusionMatrix(factor(testing$expire_flag), test_pred2, positive="yes")
```

#decision tree
```{r}
prp(dtree_fit2$finalModel, box.palette = "Reds")
```

#Random Forest
```{r}
#using random search 
trctrl_rf <- trainControl(method="repeatedcv", number=10, repeats=5, search="random")
set.seed(3233)
mtry <- sqrt(ncol(training))
tunegrid <- expand.grid(.mtry=mtry)
rf_gridsearch <- train(expire_flag ~ ., data = training, method="rf", 
                       trControl=trctrl_rf,
                       preProcess = c("center", "scale"), 
                       tuneLength = 15)
print(rf_gridsearch) 
plot(rf_gridsearch)
```

#usign rf_gridsearch to predict on testing set
```{r}
test_pred <- predict(rf_gridsearch, newdata = testing)
test_pred
```

#confusion matrix
```{r}
confusionMatrix(factor(testing$expire_flag), test_pred, positive="yes")
```

The purpose of this project is to utilize data extracted out from MIMIC III to train models and predict our desired outcome. Our team chose to investigate in premature health attacks- looking into patients who are younger than 65 that got admitted into the hospital due to heart attacks. We were interested in looking into factors such as age, icu length of stay, weight, height, cholesterol levels and many more. The first issue we faced was that once we obtained the mimic III dataset, there are lots of NAs for multiple variables. As a result, I had to remove many variables that would have played important roles in predicting patients’ mortalities such as height, weight, heart rate and cholesterol levels. For the data cleaning part, I decided to re-level the ethnicities: instead of having over 20 different levels, I concentrated it down to 5: Asian, Black, White, Hispanic and Others. I also renamed the expire_flag levels from 0 and 1 to no and yes. Upon further investigation, several variables were removed as well if it contains more than 20% NAs. A correlation plot is produced to get an idea of what variables might be strongly correlated to our predictor variable- expire_flag, and nothing really popped into significance. Data visualizations have shown that there are more males than females in the dataset, and the race of the patients is predominantly white. 
For data modeling, I tried multiple methods including logistic regression, SVM classifier with linear, poly and rbf kernel, decision tree and random forest. Overall, the results are not too great, and the performance between each method is not significantly different. Simple logistic regression actually outperformed SVM linear and SVM poly by 1%. SVM Radial performed slightly better than linear and poly kernel, but the difference is approximately 1% as well. Most models gave higher specificity than sensitivity- indicating that the true negative rate is higher than the true positive rate. At the end, random forest provided the best result with accuracy rate up to 70.8%, sensitivity 64.4% and specificity at 72.3 %. Decision tree is the second-best performer with accuracy rate of 68.3%. These models didn’t perform well in predicting patients mortality rate can be due to the fact that we were unable to include important factors such as weight, cholesterol level and heart rate. At the end, we were left with variables that are usable, but these variables might not be highly related to heart attacks. 

Logistic: 0.658
SVM Linear: 0.648
SVM poly: 0.648
SVM Radial: 0.655
Decision tree: 0.683
Random forest: 0.708
